---
title: "Discrete Probability Notes"
author: "Luis Moreno"
output: html_notebook
---
# [Probability](https://rafalab.github.io/dsbook/probability.html)  

## Discrete Probability  
- The probability distribution for a variable describes the probability of observing each possible outcome.  
- For discrete categorical variables, the probability distribution is defined by the proportions for each group.  

### Relative Frequency  

Probability of an event: the proportion of **times the event occurs** when we repeat the experiment independently under the same conditions.  
Event: an outcome that can occur when something happens by chance.  

$P(A)$ = "Probability of event A happening". 

### Probability Distributions  
Assign a probability to each category relative frequency.  

## Monte Carlo Simulations for Categorical Data  

Monte Carlo simulations model the probability of different outcomes by repeating a random process a large enough number of times that the results are similar to what would be observed if the process were repeated forever.  

- The `sample()`function let us mimic the process of picking at random.  
     * By setting the argument `replace = TRUE`, to repeat the function continually.  
- The `rep()` function let us generate random numbers.  
- The `replicate()` function, which permits us to repeat the same task any number of times.  
- The `table()` function let us see the distribution of a variable.  

```{r Sample Simulation}
# Simulate a coin toss
coins <- rep(c("head", "tails"), 
             times = c(2,3))
coins

# Pick one random value
sample(coins, 1)

events <- 
     # Replicate a function a thousand times
     replicate(1000, 
               # Pick one random value
               sample(coins, 1))

tab <- 
     # Create a table that counts each combination of factor levels
     table(events)
tab

# Return proportions of an R object
prop.table(tab)
```

- The `set.seed()` function ensures that results are exactly the same every time you run them. It is useful when dealing with random number generators.  
     * `set.seed(1, sample.kind="Rounding")` will make R 3.6 generate a seed as in R 3.5.  

- The `mean()` function is used to calculate the arithmetic mean. If applied to a logical vector, The mean of the zeros and ones thus gives the proportion of `TRUE` values.  

## Independence  
Two events are independent if the outcome of one does not affect the other.  

## Conditional Probabilities  
When events are dependent, conditional probabilities are useful.  

The following notation is useful for conditional probabilities:  

- $Pr(Card2 = King \mid Card1=King)$  

The "|" is used as shorthand for "given that" or "conditional on".  

When two events are independent, we have:  

- ${P(A \mid B)} = {Pr(A)}$ 

In other words, "the fact that B happened does not affect the probability of A happening".  

## Addition and Multiplication Rules  
### Multiplication Rule  

* Pr(A and B) = Pr(A) Pr(B∣A)  

The multiplication rule also applies to more than two events. We can use induction to expand for more events:  
* Pr(A and B and C) = Pr(A) Pr(B ∣ A) Pr(C ∣ A and B)  

### Multiplication Rule Under Independence  
When dealing with independent events, the multiplication rule becomes simpler:  

* Pr(A and B and C) = Pr(A)Pr(B)Pr(C)  

Assuming independence can result in very different and incorrect probability calculations when we do not actually have independence.  

The multiplication rule also gives us a general formula for computing conditional probabilities:  

- ${Pr(B \mid A)} = \frac {Pr(A and B)} {Pr(A)}$  

### Addition Rule  
Pr(A or B) = Pr(A) + Pr(B) − Pr(A and B)  

This rule is intuitive: think of a Venn diagram. If we simply add the probabilities, we count the intersection twice so we need to subtract one instance.  

```{r Probability and Sampling}
cyan <- 3
magenta <- 5
yellow <- 7

# The variable 'p_1' is the probability of choosing a cyan ball from the box on the first draw.
p_1 <- cyan / (cyan + magenta + yellow)

# Assign a variable 'p_2' as the probability of not choosing a cyan ball on the second draw with replacement.
p_2 <- 1 - p_1

# Calculate the probability that the first draw is cyan and the second draw is not cyan using `p_1` and `p_2`.
p_1 * p_2
```

### Combinations and Permutations  
```{r Paste and Expand Grid}

# joining strings with paste
number <- "Three"
suit <- "Hearts"
paste(number, suit)

# joining vectors element-wise with paste
paste(letters[1:5], 
      as.character(1:5))

# generating combinations of 2 vectors with expand.grid
expand.grid(pants = c("blue", "black"), 
            shirt = c("white", "grey", "plaid"))
```

```{r Generating a Deck of Cards}
suits <- c("Diamonds", "Clubs", "Hearts", "Spades")
numbers <- c("Ace", "Deuce", "Three", 
             "Four", "Five", "Six", 
             "Seven", "Eight", "Nine", 
             "Ten", "Jack", "Queen", "King")

# create a data frame from all combinations of numbers and suits
deck <- expand.grid(number = numbers, 
                    suit = suits)

# paste the values of two columns
deck <- paste(deck$number, 
              deck$suit)

# probability of drawing a king
kings <- paste("King", suits)
mean(deck %in% kings)
```

```{r Permutations and Combinations}
library(gtools)
# ways to choose 2 numbers in order from 1:5
permutations(5,2)

# ways to choose 7 numbers in order from 1:10 using digits 0:9
all_phone_numbers <- permutations(10, 7, v = 0:9)

# count the number of rows in a vector
n <- nrow(all_phone_numbers)

# take a sample of 5 elements from a vector "n"
index <- sample(n, 5)

# subset values by sampling
all_phone_numbers[index,]

permutations(3,2)    # order matters
combinations(3,2)    # order does not matter
```

```{r Conditional Probability when Order Matters}
# compute all possible ways we can choose two cards when the order matters
# ways of choosing 2 cards from 1:52 using deck as values
hands <- permutations(52,2, v = deck)

# get the first and second cards
first_card <- hands[,1]
second_card <- hands[,2]

# compute the cases for which the first hand was a King
kings <- paste("King", suits)
sum(first_card %in% kings)

# To get the probability of drawing a second king given that one king is drawn, we compute what fraction of these have a King in the second card
## Method 1
sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)
## Method 2
mean(first_card%in%kings & second_card%in%kings) / mean(first_card%in%kings)
```

```{r Conditional Probability when Order Does Not Matters}
# Compute the probability of getting an Ace and a face card in the first draw (which is known as a Natural 21) using combinations
aces <- paste("Ace", suits)
facecard <- c("King", "Queen", "Jack", "Ten")

# get all possible combinations between "facecard" and "suits"
facecard <- expand.grid(number = facecard, 
                        suit = suits)

# paste all possible combinations of "facecard" and "suits" into one vector
facecard <- paste(facecard$number, 
                  facecard$suit)

# get all possible hands
hands <- combinations(52, 2, v=deck)

# probability of a natural 21 given that the ace is listed first in `combinations`
mean(hands[,1] %in% aces & hands[,2] %in% facecard)

# probability of a natural 21 checking for both ace first and ace second
mean((hands[,1] %in% aces & hands[,2] %in% facecard)|(hands[,2] %in% aces & hands[,1] %in% facecard))
```

```{r Monte Carlo Simulation}
# code for one hand of blackjack
hand <- sample(deck, 2)
hand

# code for B=10,000 hands of blackjack
B <- 10000

results <- 
        replicate(B, {
                hand <- sample(deck, 2)
                (hand[1] %in% aces & hand[2] %in% facecard) | 
                        (hand[2] %in% aces & hand[1] %in% facecard)
                })

mean(results)
```

#### Birthday Problem  
```{r The Birthday Problem}
# set the sample size
n <- 50

# ways of choosing 50 values from 1:365 without replacement
bdays <- sample(1:365, n, replace = TRUE)

# check if at least two values are the same
any(
        # check if an element of a vector is duplicated
        duplicated(bdays)
        )
# set the number of samplings
B <- 10000

# function to calculate repeated and similar values when sampling
same_birthday <- function(n){
  
  bdays <- 
    
    # take a sample of size "n" from 1:365 with replacement
    sample(1:365, n, replace = TRUE)
  
  # check how many values are the same
  any(
    
    # check if an element is duplicated
    duplicated(bdays))
}

results <- 
        # repeat "B" times the "same_birthday" function by sampling 50 values
        replicate(B, same_birthday(50))

mean(results)

```

#### Vectorization and Functionals  

* In R, **vectorization** is preferred over for-loops since it results in shorter and clearer code.  

* A **vectorized function** is a function that will apply the same operation on each of the vectors. However, not all functions work this way.  

* **Functionals** are functions that help us apply the same function to each entry in a vector, matrix, data frame, or list.  

* The function `sapply` performs element-wise operations on any function. It operates on numeric, logical, and character vectors.  
  - Other functionals are `apply`, `lapply`, `tapply`, `mapply`, `vapply`, and `replicate`.  

```{r Element-wise Operations on Any Function}
# create a function with two arguments
compute_prob <- function(n, B = 10000){
  
  # replicate "B" times the "same_birthday" function
  results <- replicate(B, same_birthday(n))
  
  # get the arithmetic 
  mean(results)
}

# create a sequence from 1 to 60
n <- seq(1, 60)

# apply the "compute_prob" function by "n" times
prob <- sapply(n, compute_prob)

# load packages
library(tidyverse)

# use quick plot to plot the estimated probabilities of two people having the same birthday in a group of size "n"
qplot(n, prob)
```

```{r Computing and Plotting Probabilities}
# create a function with one argument
exact_prob <- function(n){
  
  # create a decreasing fraction sequence starting at 365
  # to calculate the probability of a group "n" having the same birthday
  prob_unique <- seq(365, 365 - n + 1) / 365
  
  # subtract to 1 the "product" of all values in "prob_unique",
  # to obtain the probability of and event not happening
  1 - prod(prob_unique)
}

# apply "exact_prob" to the vector "n"
eprob <- sapply(n, exact_prob)

qplot(n, prob) + 
  
  # add a red line to the plot
  geom_line(aes(n, eprob), 
            col = "red")
```

### Infinity in Practice  

In practice we can't repeat an experiment forever. The more experiments we perform, the better the approximation.  

How many experiments do we need to perform?  

```{r Estimates Stability}
# defines vector of many B values
B <- 10^seq(1, 5, len = 100)

# function to run Monte Carlo simulation with each B
compute_prob <- function(B, n = 25){
  same_day <- replicate(B, same_birthday(n))
  mean(same_day)
}

# apply compute_prob to many values of B
prob <- sapply(B, compute_prob)

# plot a line graph of estimates 
qplot(log10(B), prob, geom = "line")
```

The values start to stabilize (they vary less than .01) around 1,000. The exact probability that was previously calculated was 0.569.  

