---
title: "Week 2 Peer Review"
output: html_document
---

```{r setup, include=FALSE}

######DO NOT MODIFY. This will load required packages and data.

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
cces <- drop_na(read_csv(url("https://www.dropbox.com/s/ahmt12y39unicd2/cces_sample_coursera.csv?raw=1")))
```

Your objective is to replicate these figures, created using the Cooperative Congressional Election Study data. These figures are similar to those we completed in the lecture videos.

### Put your name here: LUIS MORENO



## Exercise 1

Hints:

* For the x-axis, use the variable "ideo5".

* Make sure you recode the data for the "ideo5" variable to generate the correct names for the x-axis. You will want to consult the codebook.

* Use the fill aesthetic to have R fill in the bars. You do not need to set the colors manually.

* Use guides() to drop the legend.

* Make sure the axis labels and figure title are correct.


```{r exercise1,echo=FALSE}
ideo5 <- recode(cces$ideo5,
                `1` = "Very Liberal",
                `2` = "Liberal",
                `3` = "Moderate",
                `4` = "Conservative",
                `5` = "Very Conservative")

cces$ideo5 <- ideo5

ggplot(cces, aes(x = ideo5, fill = ideo5)) +
     
     geom_bar() +
     
     labs(title = "Respondent Ideology", 
          x = "Ideology", 
          y = "Count") +
     
     guides(fill = FALSE)
```

## Exercise 2

Hints:

* For the x-axis, use the variable "pew_religimp".

* Make sure you recode the data for the "pew_religimp" variable to generate the correct labels for the x-axis. You will want to consult the codebook.

* Rename the column for Ideology to make sure the first letter is upper-case (to make the legend appear correctly).

* Use the fill aesthetic to have R fill in the bars. You do not need to set the colors manually.

* Make sure the axis labels and figure title are correct.

```{r exercise2,echo=FALSE}
pew_religimp <- recode(cces$pew_religimp, 
                       `1` = "Very imp.",
                       `2` = "Somewhat imp.",
                       `3` = "Not too imp.",
                       `4`= "Not imp.")

cces$pew_religimp <- pew_religimp

cces %>% 
     rename("Ideology" = "ideo5") %>%
     
     ggplot(aes(x = pew_religimp,
                fill = Ideology)) +
     
     geom_bar() +
     
     labs(title = "Importance of Religion by Ideology",
          x = "Importance of Religion",
          y = "Count")
```

## Exercise 3

Instructions:

* For this visualization, you are creating your own data for practice.

* Create a tibble/data frame with three columns: Semester, Student, and Grade.

* There should be six semesters and three students (Amanda, Betty, and Carol)

* Create grades for the students using the runif() command, with values between 80 and 100. Hint: you'll need 18 grades total.

The figure should look approximately like this (your vaules will be slightly different):


```{r exercise3,echo=FALSE}
semesters <- seq(from = 1, to = 6, by = 1)
students <- c("Amanda", "Betty", "Carol")
dat <- expand.grid(X=semesters, Y=students)
grades <- runif(18, 80, 100)

education <- tibble(dat, grades)
     education %>% rename("Semester" = "X",
                          "Student" = "Y",
                          "Grade" = "grades") %>%
          ggplot(aes(x = Semester,
                     y = Grade,
                     color = Student)) +
          geom_line(show.legend = FALSE) +
          facet_wrap(~Student) +
          labs(title = "Student Grades by Semester")
```
